---
title: "ADSP 31006 Time Series Analysis and Forecasting"
output: 
  html_document:
    code_folding: show
    theme:
      bg: "#202123"
      fg: "#B8BCC2"
      primary: "#EA80FC"
      base_font:
        google: Prompt
      heading_font:
        google: Proza Libre
      version: 3
    toc: true
    toc_float: true
    smooth_scroll: true
---

```{r setup, include=FALSE}
if (requireNamespace("thematic")) 
  thematic::thematic_rmd(font = "auto")

library(ggplot2)
library(ggfortify)
library(forecast)
```

# Assignment #1 - Introduction
## Vishal Parameswaran

### Question 1:
Suppose:

  1. $E(X) = 2$
  2. $Var(X) = 9$ 
  3. $E(Y) = 0$ 
  4. $Var(Y) = 4$ 
  5. $Corr(X,Y) = 0.25$ 

Find:

  1. $Var(X+Y)$ 
  2. $Cov(X,X+Y)$ 
  3. $Corr(X+Y,X-Y)$ 

ANS:
Considering:

  1. $E(X) = 2$
  2. $Var(X) = 9$ 
  3. $E(Y) = 0$ 
  4. $Var(Y) = 4$ 
  5. $Corr(X,Y) = 0.25$ 

  then

---

#### 1. $Var(X+Y)$
  
  We know that 
  
  $Var(X+Y) = Var(X) + Var(Y) + 2Cov(X,Y)$.
  
  To solve the equation, we need to solve $Cov(X,Y)$ first. We know that 
  
  $Corr(X, Y) = \frac{Cov(X,Y)}{\sqrt{(Var(X) * Var(Y))}}$. 
  
  Rearranging for $Cov(X,Y)$ we get 
  
  $Cov(X,Y) = Corr(X, Y)\sqrt{(Var(X) * Var(Y))}$. 

  Solving the above equation we get

  $Cov(X,Y) = 0.25*\sqrt{9*4} = 0.25*\sqrt{36} = 0.25*\pm6 = \pm1.5$

  Since the standard deviation cannot be negative, we take the positive value of 1.5. Thus
  
  $Cov(X,Y) = 1.5$

  Now we can solve for $Var(X+Y)$.

  $Var(X+Y) = Var(X) + Var(Y) + 2Cov(X,Y)$

Therefore

$Var(X+Y) = 9 + 4 + 2*1.5 = 16$

---

#### 2. $Cov(X,X+Y)$

  We know that 
  
  $Cov(X,X+Y) = Cov(X,X) + Cov(X,Y)$
  
  $Cov(X,X+Y) = Var(X) + Cov(X,Y)$

  $Cov(X,X+Y) = 9 + 1.5 = 10.5$

  Thus $Cov(X,X+Y) = 10.5$

---

#### 3. $Corr(X+Y,X-Y)$

  We know that

  $Corr(X+Y,X-Y) = \frac{Cov(X+Y,X-Y)}{\sqrt{(Var(X+Y) * Var(X-Y))}}$

  $Var(X+Y) = 16$ (from above)

  $Var(X-Y) = Var(X) + Var(Y) - 2Cov(X,Y)$

  $Var(X-Y) = 9 + 4 - 2*1.5 = 12 - 3 = 10$

  $Cov(X+Y,X-Y) = Cov(X,X-Y) + Cov(Y,X-Y)$

  $Cov(X+Y,X-Y) = Cov(X,X) - Cov(X,Y) + Cov(Y,X) - Cov(Y,Y)$

  $Cov(X+Y,X-Y) = Var(X) - Cov(X,Y) + Cov(Y,X) - Var(Y)$

  $Cov(X+Y,X-Y) = 9 - 1.5 + 1.5 - 4 = 9 - 4 = 5$

  $Corr(X+Y,X-Y) = \frac{5}{\sqrt{(16 * 10)}} = \frac{5}{\sqrt{160}}$

  $Corr(X+Y,X-Y) = 0.3953$

---

### Question 2:
If $X$ and $Y$ are dependent but $Var(X) = Var(Y)$ Find $Cov(X+Y,X-Y)$

Ans:

---

  We know that $X$ and $Y$ are dependent and $Var(X) = Var(Y)$

  $Cov(X+Y,X-Y) = Cov(X,X-Y) + Cov(Y,X-Y)$

  $Cov(X+Y,X-Y) = Cov(X,X) - Cov(X,Y) + Cov(Y,X) - Cov(Y,Y)$

  $Cov(X+Y,X-Y) = Var(X) - Cov(X,Y) + Cov(Y,X) - Var(Y)$

  substituting $Var(X) = Var(Y)$ and $Cov(X,Y) = Cov(Y,X)$
  
  $Cov(X+Y,X-Y) = Var(X) - Cov(X,Y) + Cov(X,Y) - Var(X)$

  $Cov(X+Y,X-Y) = Var(X) - Var(X)$

  $Cov(X+Y,X-Y) = 0$

---


### Question 3:
Suppose $Y_{t} = 5 + 2t + X_{t}$ where $X_{t}$ is a zero-mean stationary series with autocovariance function $\gamma_{k}$.

a. Find the mean function for $\{Y_{t}\}.$
b. Find the autocovariance function for $\{Y_{t}\}.$
c. Is $\{Y_{t}\}$ stationary? Why or why not?

Ans:

---

  We know that $Y_{t} = 5 + 2t + X_{t}$ where $X_{t}$ is a zero-mean stationary series with autocovariance function $\gamma_{k}$.
  
  Therefore $E(X_{t}) = 0$

  The mean function can be calculated as follows:

  $\mu_{Y} = E(Y) = \int_{-\infty}^{\infty}Yf(Y)dY$
  
  $\mu_{Y} = E(5 + 2t + X_{t}) = 5 + 2t + E(X_{t}) = 5 + 2t + 0 = 5 + 2t$

  Which is not constant over time as it is dependent on time $t$.

---

  The autocovariance function of $Y_{t}$ over time $k$ can be calculated as follows:
  $Cov(Y_{t}, Y_{t-k}) = Cov(5 + 2t + X_{t}, 5 + 2(t-k) + X_{t-k})$ :  Expanding on $Y_{t}$

  $Cov(5 + 2t + X_{t}, 5 + 2(t-k) + X_{t-k}) = Cov(X_{t}, X_{t-k})$ : Covariance of constants is 0

  $Cov(X_{t}, X_{t-k}) = \gamma_{k}$

  Therefore the autocovariance function of $Y_{t}$ over time $k$ is $\gamma_{k}$


---


  $\{Y_{t}\}$ is not stationary because the mean function $E(Y_{T}) = 5 + 2t$ is not constant over time.

---

### Question 4:
The following time plots and ACF plots correspond to three different time series. Your task is to match each time plot in the first row with one of the ACF plots in the second row.

ANS:

---

  I would match 
  
  (1) monthly accidental rate (1) with ACF plot (a) as the time series plot shows a very consistent seasonal pattern with a period of 12 months. The ACF plot shows the seasonality in its sticky nature.

  (2) monthly air passengers (2) with (c) as the time series plot shows a very clear trend line, which is displayed in the ACF plot with a very slow decay curve to 0.

  (3) Annual mink trappings (3) with (b) as the time series plot shows a not so clear seasonal pattern, which displays some deviation from the seasonal movements. The ACF plot shows this with the lines not being very sticky, and also changing direction abruptly.

---


### Question 5:
Simulate a normal white noise sample of size 10 with the mean 2.3 and standard deviation 1.2, and then 
calculate the sample mean and sample standard deviation. Are you satisfied with your computed
results? How about the sample size 10,000?

Ans:

---

Lets simulate a white noise sample of size 10 with the mean 2.3 and standard deviation 1.2. We know what to satisfy the conditions of being white noise

1.  for all $t$, $E(W_{t}) = μ$ is a constant;
2.  for all $t$, $Var(W_{t}) = \sigma^{2}$ ; w is a constant
3.  it is uncorrelated at different time points, that is, when $t = s$, $Cov(W_{t}, W_{s}) = 0$.

```{r, fig.showtext=TRUE}
set.seed(30)
size <- 10
mean_data <- 2.3
sd_data <- 1.2
y <- ts(rnorm(size, mean_data, sd_data))
autoplot(y) + ggtitle("White Noise Sample of Size 10") + xlab("Time") + ylab("y") # nolint
```

The sample mean and sample standard deviation can be calculated as follows:
  
```{r, fig.showtext=TRUE}
sample_mean <- mean(y)
sample_sd <- sd(y)
```
```{r, fig.showtext=TRUE}
sample_mean
sample_sd
```

The sample does not seem to show white noise due to the lack of plot points in the data. Thus the sample mean is unable to reach the population mean. Its the same when we consider the standard deviation as well. Maybe simulating more points will help.

```{r, fig.showtext=TRUE}
set.seed(30)
size <- 10000
mean <- 2.3
sd <- 1.2
y <- ts(rnorm(size, mean, sd))
autoplot(y) + ggtitle("White Noise Sample of Size 10000") + xlab("Time") + ylab("y") # nolint
```
```{r, fig.showtext=TRUE}
sample_mean <- mean(y)
sample_sd <- sd(y)
```
```{r, fig.showtext=TRUE}
sample_mean
sample_sd
```

With a larger number of plot points, we are able to see a more accurate representation of white noise. The sample mean is closer to the population mean and the sample deviation is closer to the population standard deviation. 

---

### Question 6:
R comes with two time series that are already R's `ts()` class: AirPassengers and JohnsonJohnson. 

a. What is each of those time series reporting?
b. Find the start date, the end date and frequency of each of the time series. 
c. Plot separate time plots of the two time series. Describe what the time plots tell us about the time series and compare the two plots. What are the main features of each of the time series?
d. Plot the seasonal box plots of the two time series and describe what those plots are conveying 
about the time series.

Ans:

---

a.  The `AirPassengers` time series reports the monthly totals of international airline passengers from 1949 to 1960. 

The `JohnsonJohnson` time series reports the quarterly earnings per Johnson & Johnson share from 1960 to 1980.

---

b. The start date, end date and frequency of each of the time series can be found as follows:

#### AirPassengers
```{r, fig.showtext=TRUE}
ts(AirPassengers)
start(AirPassengers)
end(AirPassengers)
frequency(AirPassengers)
```
#### JohnsonJohnson
```{r, fig.showtext=TRUE}
ts(JohnsonJohnson)
start(JohnsonJohnson)
end(JohnsonJohnson)
frequency(JohnsonJohnson)
```

---

c. The time plots of the two time series can be plotted as follows:

```{r, fig.showtext=TRUE}
autoplot(AirPassengers) + ggtitle("AirPassengers") + xlab("Time") + ylab("Passengers") # nolint
```

```{r, fig.showtext=TRUE}
autoplot(JohnsonJohnson) + ggtitle("JohnsonJohnson") + xlab("Time") + ylab("Earnings") # nolint

```

From the above plot we can clearly see that seasonality and trend are present in the data. The trend is increasing over time and the seasonality is sticky. The data is not stationary as the mean is not constant over time. The  `AirPassengers` dataset has more spiky seasonality than the `JohnsonJohnson` dataset. The `AirPassengers` dataset has a more linear trend than the `JohnsonJohnson` dataset. Moreover the `JohnsonJohnson` dataset does not start out with a clear seasonality, but develops it over time as the data progresses.



---

d. The seasonal box plots of the two time series can be plotted as follows:

```{r, fig.showtext=TRUE}
season <- cycle(AirPassengers)
season.factor <- factor(season)
ggplot(AirPassengers, aes(x = season.factor,
                             y = AirPassengers,
                             fill = season.factor)) +
  geom_boxplot() +
  labs(x = "Month", y =  "Passengers")
```

```{r, fig.showtext=TRUE}
seasonJ <- cycle(JohnsonJohnson)
seasonJ.factor <- factor(seasonJ)
ggplot(JohnsonJohnson, aes(x = seasonJ.factor,
                             y = JohnsonJohnson,
                             fill = seasonJ.factor)) +
  geom_boxplot() +
  labs(x = "Month", y =  "Earnings")
``` 

Looking at the seasonal box plots, we can see that seasonality exists, as the means for each season are clustered towards the middle of each season. The `AirPassengers` dataset shows a greater degree of  seasonality as there is a clear variation across seasons. In the `JohnsonJohnson` dataset, the seasonality is not as clear as the frequency is lower. Looking at both the Time Series plot and the seasonality box plot for the `JohnsonJohnson` dataset we can see that the seasonality is across a year and not a quarter. This is because the data is quarterly and not monthly.

---
